\blue{Numerical Investigation of Multi-Fluid Flows through Dual Saturated-Unsaturated Porous Media - change the title to something close to the main topic of the manuscript ... something with viscous flow instability or viscous fingering)}

### 2017-02-01 Papers/ViscousInstabilities/Manuscript_ViscousInstabilities_KC.tex
updates

Numerical investigation and modelling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations. Out of this mixed and complex formations, viscous flow instabilities can occur both within a single phase or when multiple phases flow through porous media. Forecasting the behaviour of multiphase flows or the future response of the system are major challenges that need to be taken into account. Injections of low-viscosity, high-density fluid $1$ (e.g. water) into a reservoir saturated with lighter but more viscous oil is an example where both viscous and gravitational instabilities may occur for two-phase immiscible and incompressible flow. This is mainly due to the importance of the oil and gas recovery operations, groundwater exploration, underground coal gasification, and more recently on CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture and storage (CCS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

\medskip

The conservation of mass along with Darcy's law are the governing equations that describe the flow pattern. Particle methods, front-tracking methods \citep{univerdi_1992} but in particular volume of fluid (VOF) methods \citep{hirt_1981,rider_1998} are used to describe the afformentioned proeprties. In addition, finite difference methods have been applied to discretise the governing equations \citep{Luo2016,Moortgat2016,Hoteit2008} .

The objective of this work is to investigate and accurately describe the behaviour of fluid instabilities in compositional flows using a control volume finite element method (CVFEM). Considering the limitations of finite differences formulations the CVFE The model formulation will use a conservative computational multi-fluid porous media flow model able to exploit mesh adaptivity methods on a fully-unstructured tetrahedral grid. Also two key characteristics that describe the model above are: (a) the hybrid element pair $P_{n}DGP_{n+1}$ which provides discontinuous and piecewise representation for velocity and pressure fields at FE space and (b) the overlapping control volume finite element method (CVFEM)formulation.

Larger scale problems can be approached by upscaling techniques. These are techniques which are usually employed to solve problems that fail because of certain types of heterogeneities. These cases are normally resolved by stochastic representation of permeability and porosity fields. There are several approaches to upscale solutions, while three common types are discussed. The $1^{st}$ assumes capillary equilibrium and calculates effective relative permeability using a suitable technique, such as numerical methods, percolation methods and space re-normalization methods. The $2^{nd}$ approach is called pseudoization. It generates a psuedo relative permeability by matching immiscible displacement simulation of a coarse-grid to the output of fine-grid simulations. The $3^{rd}$ approach involves direct calculation of relative permeability from displacement experiments.~\citep{johnson_1959, jones_1978}. Problems start to rise when complex flow patterns will appear during the recovery of hydrocarbons triggered by heterogeneities.  
%The first approach is limited to low flow rates and very small heterogeneities. The second approach overcomes these limitations but also posses drawbacks such us, unavailability of fine-grid simulation with suitable level of accuracy, dependence of the pseudo functions on rate, direction and grid specification. The experimental data from the third approach is only suitable for a small class of scenarios.~\citep{Chang_1994}. While  non-uniform permeability and porosity may be easy to deal with at a microscopic scale for porous media flow equations, it presents a lot of challenges at a meso/field scale, which is required for reservoir simulations.

@@

\medskip
Under certain geological conditions the displacement of fluids in porous media is unstable and results in the formation of irregular flow patterns in the form of fingers %\textit{fingers}. 
These patterns are also described as flow instabilities or \textit{viscous fingering},  or most widely known as Saffman-Taylor instabilitises. This phenomenon will occur during an unfavorable -mobility-ration displacement, when a more viscous fluid is displaced by a less viscous (Saffman and Taylor $1958$). The mechanism of the instabilities may vary, 
%(in immiscible fluids the most important factor is the surface tension while in miscible fluids, it is the diffusion \red{(thermal or mass diffusion ...be specific)})
however it is still a function of relative mobility, gravity, viscous forces and permeability differences. In conventional reservoirs the principal forces that affect fluid flow include, viscous forces, effects of gravity and permeability diffrences while in unconventional gas reservoirs flow simulation models take into account the phenomena of adsorption and diffusion \citep{abdus_2015}.
Fingering will occur in miscible displacments when the viscocity ratio is greater than unity. As surface tension becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability by the basic mechanism associated with a less viscous fluid displacing a more viscous one. After a split, each of the new lobes of the finger is stable as a result of their being thinner than the finger from which they split. As a result of \textit{shielding}, one of these lobes will eventually outgrow the others and will then spread to occupy the appropriate width of the cell. In the process, it reaches a width that is again unstable to splitting, and the pattern repeats. Thus, surface tension plays an essential dual role; it must be weak enough for the tip front to be unstable, but it is also the physical force causing the \textit{spreading} and ensuing repeated branching. Important parameter is the mobility (or viscosity) ratio. This is given by Eqn.~\ref{eq:mobility-of-fluid} where $K$ is absolute permeability and $k_{r}$ is relative permeability :

@@

% are based on domains that require high resolution static meshes to capture the development of the fingers 
Even though some commercial simulators provide the user with option of an unstructured type of mesh still they are not able to dynamically adapt the mesh as the simulation develops and they increase the computational cost as the saturation front changes slowly. In this paper the use of dynamic mesh adaptivity to capture the viscous fingering in reservoir flows is introduced, as one of the methods that will allow the user to focus the effort where is needed, minimizing both the CPU time and the numerical diffusion.

The remainder of the paper is orgranised as follows.

\KCnote{ this needs to be rephrased later -- Next section \citet{math-model} illustrates the model and the governing equations of immiscible and incompressible flow in porous media. The method and numerical formulations that describes the CVFEM follows, along with cases that demonstrate how the mesh adaptivity for simulation of viscous fingering is implemented. In the end, conclusions and final remarks of the work are discussed.--}

@@

@@

\KCnote{need to include some figures and analysis of the velocities, magnitude etc.}

>>>>

Numerical investigation and modelling of multiphase flows in porous media have attracted the attention of the scientific community over the past 40 years. Characterization of the flow of fluids through porous media serves as the foundation of reservoir engineering studies. Reservoir engineers and hydro-geologists are dealing with complex geometries, irregular boundaries and heterogeneous geological formations. Out of this mixed and complex formations, viscous flow instabilities can occur both within a single phase or when multiple phases flow through porous media. Forecasting the behaviour of multiphase flows or the future response of the system are major challenges that need to be taken into account. Injections of low-viscosity, high-density fluid $1$ (e.g. water) into a reservoir saturated with lighter but more viscous oil is an example where both viscous and gravitational instabilities may occur for two-phase immiscible and incompressible flow. These phenomena have great importance during the oil and gas recovery operations in both mature and new fields. Starting from oil and gas migration of contaminants through groundwater is the opposite problem governed by the same principles. The most complicated of processes include fingering that wil loccur in multiphase and compositional flow. Underground coal gasification is another potential application but more recently on CO$_{\text{2}}$ migration and trapping mechanisms in carbon capture and storage (CCS) operations~\citep{spycher_2003, chen_2006, self_2012, pruess_1990c, white_1981,jiang_2011}.

\medskip
The conservation of mass along with Darcy's law are the governing equations that describe the flow pattern. Particle methods, front-tracking methods \citep{univerdi_1992} but in particular volume of fluid (VOF) methods \citep{hirt_1981,rider_1998} are used to describe the afformentioned properties. In addition, finite difference methods have been applied to discretise the governing equations \citep{Luo2016,Moortgat2016,Hoteit2008}. This work will mainly focus on macroscale flow instabilities or \textit{viscous fingering} in heterogeneous resevoirs -or computational domains- and will compare the development of fingers using an alternative finite element method. 
In conventional reservoirs the principal forces that affect fluid flow include, viscous forces, effects of gravity and permeability diffrences while in unconventional gas reservoirs flow simulation models take into account the phenomena of adsorption and diffusion \citep{abdus_2015}. Even though the mechanisms of the instabilities may vary, however it is still a function of relative mobility, gravity, viscous forces and permeability differences. All these parameters are represented under one term in the global mass balance equation.

\medskip
The objective of this study is to investigate and accurately describe the behaviour of fingering in multiphase ($2$-phase) flows using a control volume finite element method (CVFEM). Considering the limitations of finite differences formulations the CVFE the model formulation will use a conservative computational multi-fluid porous media flow model able to exploit mesh adaptivity methods on a fully-unstructured tetrahedral grid. Also two key characteristics that describe the model above are: (a) the hybrid element pair $P_{n}DGP_{n+1}$ which provides discontinuous and piecewise representation for velocity and pressure fields at FE space and (b) the overlapping control volume finite element method (CVFEM)formulation.

\medskip
The sections that follow include a summary of the main governing equations for multiphase compositional flow. The remainder of the paper will include a description of the viscous fingering along with the finite element method technique, a discussion of simulations results and the key conclusions. The analysis will consider ($1$)permeability heterogeities, ($2$)capturing the development of the interface and the evolution of fingers, ($3$)test-case comparison, but for now will neglect any gravitational effects.

%These are techniques which are usually employed to solve problems that fail because of certain types of heterogeneities. These cases are normally resolved by stochastic representation of permeability and porosity fields. There are several approaches to upscale solutions, while three common types are discussed. The $1^{st}$ assumes capillary equilibrium and calculates effective relative permeability using a suitable technique, such as numerical methods, percolation methods and space re-normalization methods. The $2^{nd}$ approach is called pseudoization. It generates a psuedo relative permeability by matching immiscible displacement simulation of a coarse-grid to the output of fine-grid simulations. The $3^{rd}$ approach involves direct calculation of relative permeability from displacement experiments.~\citep{johnson_1959, jones_1978}. Problems start to rise when complex flow patterns will appear during the recovery of hydrocarbons triggered by heterogeneities.  
%The first approach is limited to low flow rates and very small heterogeneities. The second approach overcomes these limitations but also posses drawbacks such us, unavailability of fine-grid simulation with suitable level of accuracy, dependence of the pseudo functions on rate, direction and grid specification. The experimental data from the third approach is only suitable for a small class of scenarios.~\citep{Chang_1994}. While  non-uniform permeability and porosity may be easy to deal with at a microscopic scale for porous media flow equations, it presents a lot of challenges at a meso/field scale, which is required for reservoir simulations.

@@

%\medskip
%Under certain geological conditions the displacement of fluids in porous media is unstable and results in the formation of irregular flow patterns in the form of fingers %\textit{fingers}. 
%These patterns are also described as flow instabilities or \textit{viscous fingering},  or most widely known as Saffman-Taylor instabilitises. This phenomenon will occur during an unfavorable -mobility-ration displacement, when a more viscous fluid is displaced by a less viscous (Saffman and Taylor $1958$). The mechanism of the instabilities may vary, 
%(in immiscible fluids the most important factor is the surface tension while in miscible fluids, it is the diffusion \red{(thermal or mass diffusion ...be specific)})
%however it is still a function of relative mobility, gravity, viscous forces and permeability differences. In conventional reservoirs the principal forces that affect fluid flow include, viscous forces, effects of gravity and permeability diffrences while in unconventional gas reservoirs flow simulation models take into account the phenomena of adsorption and diffusion \citep{abdus_2015}.

\KCnote{ - - to be putted in the mathematical formulation as intro - - 
Fingering will occur in miscible displacments when the viscocity ratio is greater than unity. As surface tension becomes weak, the front of the steady finger is susceptible to a viscous-fingering instability by the basic mechanism associated with a less viscous fluid displacing a more viscous one. After a split, each of the new lobes of the finger is stable as a result of their being thinner than the finger from which they split. As a result of \textit{shielding}, one of these lobes will eventually outgrow the others and will then spread to occupy the appropriate width of the cell. In the process, it reaches a width that is again unstable to splitting, and the pattern repeats. Thus, surface tension plays an essential dual role; it must be weak enough for the tip front to be unstable, but it is also the physical force causing the \textit{spreading} and ensuing repeated branching. Important parameter is the mobility (or viscosity) ratio. This is given by Eqn.~\ref{eq:mobility-of-fluid} where $K$ is absolute permeability and $k_{r}$ is relative permeability :

@@

}

@@

\KCnote{need to include some figures and analysis of the velocities, magnitude etc.}
\KCnote{ -to be placed in results-
 Even though some commercial simulators provide the user with option of an unstructured type of mesh still they are not able to dynamically adapt the mesh as the simulation develops and they increase the computational cost as the saturation front changes slowly. In this paper the use of dynamic mesh adaptivity to capture the viscous fingering in reservoir flows is introduced, as one of the methods that will allow the user to focus the effort where is needed, minimizing both the CPU time and the numerical diffusion. Both the evolution of visualisation techniques through the extensive use of parallel computing as well as the advances in modelling and simulation of porous media flows may help to address the importance of this phenomenon which can be found in petroleum, chemical and environmental operations.}

@@
